An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-03-07 cal_demo/ms.tex
cleaning up unused text/code
Changes to make arxiv submission work

% Basic setup. Most papers should leave these options alone.
\documentclass[a4paper,fleqn,usenatbib]{../mnras}

\usepackage{newtxtext,newtxmath}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[EPICal]{An efficient feedback calibration algorithm for direct imaging radio telescopes
%The E-field Parallel Imaging Calibration Algorithm for Next-Generation Radio Telescopes
}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/MOFF_flowchart.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered electric fields are collected by the antenna elements then frequency transformed in the F-engine. The calibration is applied, then the electric field measurements are gridded using the antenna voltage patterns. The imager performs a spatial 2D FFT on the gridded data before squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

%%%%%%
%%%%%%

%We begin by establishing the mathematical framework for the calibration problem. We derive the calibration solutions for the MOFF algorithm (adopting the notation of \citealt{thy15c}), but note the result is easily extended to FFT correlator algorithms by removing the gridding step.
%
%The electric field incident on the ground, $\Er{}(\mathbf{r},f,t)$, is related to the sky electric field, $E(\hat{\mathbf{s}},f,t)$, through a Fourier transform.
%\begin{equation}
%\Er{}(\mathbf{r},f,t) = \int E(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r}\cdot \hat{\mathbf{s}}}\, \dif^2 \hat{\mathbf{s}}
%\end{equation}
%Here $\hat{\mathbf{s}}$ denotes the sine-projected unit vector for the sky angle, $\mathbf{r}$ is the observer's location (measured in wavelengths relative to an arbitrary origin), and $f$ and $t$ denote the frequency and time dependence, respectively. We will encounter several quantities which we attempt to estimate. We distinguish the ``true" values with a superscript $T$, while the estimates are denoted with a prime. We define the true antenna signal as a convolution of the antenna voltage pattern, $\beamr$, with the electric field on the ground.
%\begin{equation}
%\Er{a}^T(f,t) \equiv \int \beamr_a(\mathbf{r}-\ra) \Er{}(\mathbf{r},f,t) \, \dif^2 \mathbf{r}
%\end{equation}
%The subscript $a$ labels the antenna, and $\mathbf{r}_a$ is the location of antenna $a$.
%
%We next model the measured, uncalibrated electric field as a multiplicative complex gain and an additive noise term applied to the true antenna electric field. 
%\begin{equation}\label{eq:apply_gain}
%\Er{a}(f,t) = g^T_a(f,t) \Er{a}^T(f,t) + \widetilde{n}_a(f,t)
%\end{equation}
%Note that this quantity is neither a true or estimated value, but rather the raw electric field signal corrupted by the complex gain before we attempt to estimate the true value. The noise term is strictly receiver noise -- noise introduced by the instrument. Any sky noise is implicitly included in the time dependence of the sky electric field. As the noise of modern low frequency arrays is heavily dominated by sky noise, we will neglect $\widetilde{n}_a$ for now, but will inspect its effects at the end of this section.
%
%The goal of our calibration algorithm will be to estimate the antenna gains, $g_a$. We will assume the gains have no time dependence within the timescale of finding our solutions. Furthermore, we will treat each frequency channel independently, and drop the $f$ to simplify notation.
%
%The MOFF algorithm next calls for a calibration. We will assume we have formed an estimate of the gains after $n$ iterations of a calibration loop. We then correct the incoming electric field data stream using this current estimate.
%\begin{equation}
%\Erest{a} = \Er{a}/ g^{(n)}_a,
%\end{equation}
%These corrected fields will be used in the EPIC pipeline to derive an updated estimate for the ($n+1$)$^{\mathrm{th}}$ iteration. 
%
%A dirty image is next formed by gridding the calibrated fields with the antenna voltage pattern, Fourier transforming, squaring, and averaging in time. The estimated value for a pixel, $\s_i$, can be expressed as
%\begin{equation}
%I'(\s_i) = \left<\left| \frac{1}{\Nant}\sum_i e^{2\pi i \mathbf{r}\cdot \s_i} \sum_a \beamr_a(\ri - \ra) \frac{g^T_a}{g^{(n)}_a} \Er{a}^T(t) \right|^2 \right>_t.
%\end{equation}
%This is the final output of the MOFF correlator. However, for calibration purposes we will be interested in the electric field image just prior to squaring and averaging.
%\begin{equation}
%\Ethetaest (\s_i,t) = \frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i} \sum_a \beamr_a(\ri - \ra) \frac{g^T_a}{g^{(n)}_a} \Er{a}^T(t)
%\end{equation}
%We can simplify this expression by exchanging the sums to transform the beam term into sky coordinates.
%\begin{align}\label{eq:epix}
%\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a \frac{g^T_a}{g^{(n)}_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\
%  & \qquad\times \sum_i \beamr_b(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
%& = \frac{1}{\Nant} \sum_a \frac{ g^T_a}{g^{(n)}_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
%\end{align}

%%%%
%%%%
As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it was suggested to form a correlation of the uncalibrated antenna measurements with an image pixel from the output of the correlator. This is a statistically stationary quantity, which can be related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum needed to calibrate a simple sky of a single point source. However, we aim for a more generalized solution for arbitrarily complex sky models. We therefore study the full expression for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<\Er{a}(t) E'^*(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ calibration loop, and \spix\, is the pixel center nearest a bright calibrator of interest. The following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which contains a bright source to achieve a high signal to noise.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients (equation~\ref{eq:Cna}). These coefficients are then shipped off-chip to apply equations~\ref{eq:cal_solution}~and~\ref{eq:damping}, and any other processing to arrive at an updated estimate of the gains, $g^{(n)}_a$. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip".}
\label{fig:schematic}
\end{center}
\end{figure}

@@

%%%%%
%%%%%

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_sim_phase.pdf}
\caption{Phase error of gain estimates as a function of iteration for simulated calibration. The gains were initialized with random phases, but the calibration loop was able to recover the correct phases after about 15 iterations. Each line represents an antenna in our 51 MWA antenna sample.
}
\label{fig:sim_phase}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_sim_amps.pdf}
\caption{Gain estimate amplitudes as a function of iteration for simulated calibration. Again, each line represents an antenna in the 51 MWA antenna sample. The gain estimates were initialized randomly, while the true values were unity. After about 15 iterations we see the calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_sim_image_before.pdf}
\includegraphics[width=\columnwidth]{figures/cal_paper_sim_image_after.pdf}
\caption{Images formed during simulated calibration. \emph{Top:} An image generated from a single frequency channel and 10 ms integration after the gain estimates are randomized. As expected with random phases, the image is completely noisy, with the shape of the primary beam evident. \emph{Bottom:} An image formed after calibration, again with a single frequency channel and 10 ms integration. The ten simulated (and modeled) point sources are easily visible, and highlighted with red circles.
}
\label{fig:sim_images}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_data_phases.pdf}
\caption{Gain phase solutions as a function of calibration iteration for an LWA TBN observation. The gain estimates are initialized with zero phase, but quickly span a $2\pi$ range, and settle into relatively flat, albeit noisy, solutions. The majority of phases congregate near zero, which is not surprising given the fairly good quality image produced from uncalibrated data.
}
\label{fig:data_phase}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_data_amps.pdf}
\caption{Gain amplitudes as a function of calibration iteration for an LWA TBN observation. The gains estimates were initialized with amplitude 0.25 after inspection of the electric field values compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_before.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_after.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_gsm_beam_weighted.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images were produces with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and Cas~A sources. The galactic plane is also substantially more evident after calibration. For reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA primary beam in the right panel. While much of the GSM is visible in the calibrated image, we note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/cal_paper_errors_complex_sky.pdf}
\caption{Gain estimate errors as a function of integration time, receiver noise (line color), and calibration method. The solid lines represent the error on the EPICal derived gain estimates, while the dashed lines represent the error on the visibility based estimates. The thicker lines were derived using a full sky model in the calibration loops, and the thinner lines used only a single point source model. In the full sky model case we see all methods trend down with longer integration time as expected. For a given integration time, EPICal derived errors are on average 23\% higher than those of visibility based calibration. In the case of a single source model, the errors bottom out due to the flux in the measurements that is not modeled. In this regime the EPICal method performs slightly better than the visibility-based method due to effectively beamforming to the sky pixel where the incomplete model is most accurate. On average EPICal errors are 5\% lower in this regime.
}
\label{fig:errors}
\end{center}
\end{figure}

@@

\bibliographystyle{../mnras}
\bibliography{../epic} % if your bibtex file is called example.bib

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

%\appendix

%\section{Some extra material}

%If you want to present additional material which would interrupt the flow of the main paper,
%it can be placed in an Appendix which appears after the list of references.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Basic setup. Most papers should leave these options alone.
\documentclass[a4paper,fleqn,usenatbib]{mnras}

%\usepackage{newtxtext,newtxmath}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[EPICal]{An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes
%The E-field Parallel Imaging Calibration Algorithm for Next-Generation Radio Telescopes
}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{MOFF_flowchart.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered electric fields are collected by the antenna elements then frequency transformed in the F-engine. The calibration is applied, then the electric field measurements are gridded using the antenna voltage patterns. The imager performs a spatial 2D FFT on the gridded data before squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it was suggested to form a correlation of the uncalibrated antenna measurements with an image pixel from the output of the correlator. This is a statistically stationary quantity, which can be related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum needed to calibrate a simple sky of a single point source. However, we aim for a more generalized solution for arbitrarily complex sky models. We therefore study the full expression for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<\Er{a}(t) E'^*(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ calibration loop, and \spix\, is the pixel center nearest a bright calibrator of interest. The following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which contains a bright source to achieve a high signal to noise.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients (equation~\ref{eq:Cna}). These coefficients are then shipped off-chip to apply equations~\ref{eq:cal_solution}~and~\ref{eq:damping}, and any other processing to arrive at an updated estimate of the gains, $g^{(n)}_a$. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip".}
\label{fig:schematic}
\end{center}
\end{figure}

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_sim_phase.pdf}
\caption{Phase error of gain estimates as a function of iteration for simulated calibration. The gains were initialized with random phases, but the calibration loop was able to recover the correct phases after about 15 iterations. Each line represents an antenna in our 51 MWA antenna sample.
}
\label{fig:sim_phase}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_sim_amps.pdf}
\caption{Gain estimate amplitudes as a function of iteration for simulated calibration. Again, each line represents an antenna in the 51 MWA antenna sample. The gain estimates were initialized randomly, while the true values were unity. After about 15 iterations we see the calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_sim_image_before.pdf}
\includegraphics[width=\columnwidth]{cal_paper_sim_image_after.pdf}
\caption{Images formed during simulated calibration. \emph{Top:} An image generated from a single frequency channel and 10 ms integration after the gain estimates are randomized. As expected with random phases, the image is completely noisy, with the shape of the primary beam evident. \emph{Bottom:} An image formed after calibration, again with a single frequency channel and 10 ms integration. The ten simulated (and modeled) point sources are easily visible, and highlighted with red circles.
}
\label{fig:sim_images}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_data_phases.pdf}
\caption{Gain phase solutions as a function of calibration iteration for an LWA TBN observation. The gain estimates are initialized with zero phase, but quickly span a $2\pi$ range, and settle into relatively flat, albeit noisy, solutions. The majority of phases congregate near zero, which is not surprising given the fairly good quality image produced from uncalibrated data.
}
\label{fig:data_phase}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_data_amps.pdf}
\caption{Gain amplitudes as a function of calibration iteration for an LWA TBN observation. The gains estimates were initialized with amplitude 0.25 after inspection of the electric field values compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{cal_paper_data_image_before.pdf}
\includegraphics[width=0.3\linewidth]{cal_paper_data_image_after.pdf}
\includegraphics[width=0.3\linewidth]{cal_paper_gsm_beam_weighted.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images were produces with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and Cas~A sources. The galactic plane is also substantially more evident after calibration. For reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA primary beam in the right panel. While much of the GSM is visible in the calibrated image, we note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{cal_paper_errors_complex_sky.pdf}
\caption{Gain estimate errors as a function of integration time, receiver noise (line color), and calibration method. The solid lines represent the error on the EPICal derived gain estimates, while the dashed lines represent the error on the visibility based estimates. The thicker lines were derived using a full sky model in the calibration loops, and the thinner lines used only a single point source model. In the full sky model case we see all methods trend down with longer integration time as expected. For a given integration time, EPICal derived errors are on average 23\% higher than those of visibility based calibration. In the case of a single source model, the errors bottom out due to the flux in the measurements that is not modeled. In this regime the EPICal method performs slightly better than the visibility-based method due to effectively beamforming to the sky pixel where the incomplete model is most accurate. On average EPICal errors are 5\% lower in this regime.
}
\label{fig:errors}
\end{center}
\end{figure}

@@

\bibliographystyle{mnras}
%\bibliography{epic} % if your bibtex file is called example.bib
\input{ms.bbl}
